Generating Random Logic Programs Using Constraint Programming

### 2020-01-17 paper/paper.tex
Another redundant constraint
Cleanup: step 2
Cleanup: step 4
Fix a bug in how dependencies are treated

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}

@@

\newcommand{\variable}[1]{\texttt{\textup{#1}}}
\newcommand{\arrayd}[3]{\variable{{#1}[}{#2}\variable{]} \in {#3}}
\newcommand{\arrayt}[3]{\variable{{#3}[}{#2}\variable{] {#1}}}
\newcommand{\predicates}{\mathcal{P}}
\newcommand{\variables}{\mathcal{V}}
\newcommand{\constants}{\mathcal{C}}
\newcommand{\tokens}{\mathcal{T}}
\newcommand{\arities}{\mathcal{A}}
\newcommand{\maxArity}{\mathcal{M}_{\mathcal{A}}}
\newcommand{\maxNumNodes}{\mathcal{M}_{\mathcal{N}}}
\newcommand{\maxNumClauses}{\mathcal{M}_{\mathcal{C}}}

@@

@@

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$} \times
    \getDependencies{$p_2$}$ such that $d_1.\predicate = d_2.\predicate$}{
    \lIf{$d_1$.\isDetermined{} {\bf and} $d_2$.\isDetermined{}}{\fail{}}
    \uIf{$d_1$.\isDetermined{}}{
      $\mathbf{A}[d_2.\source][d_2.\target]$.\removeValue{$1$}\;
    }
    \ElseIf{$d_2$.\isDetermined{}}{
      $\mathbf{A}[d_1.\source][d_1.\target]$.\removeValue{$1$}\;
    }
  }
  \caption{Propagation}
\end{algorithm}

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$} \times
  \getDependencies{$p_2$} \mid d_1.\predicate = d_2.\predicate \}$\;
  \If{$\exists (d_1, d_2) \in D$ such that $d_1.\isDetermined{}$ {\bf and}
    $d_2.\isDetermined{}$}{
    \Return{\textsc{false}}\;
  }
  \lIf{$D = \emptyset$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment}
\end{algorithm}

\begin{algorithm}[p]
  \SetKwData{edgeExists}{edgeExists}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{an $n \times n$ adjacency matrix $\mathbf{A}$}
  \Fn{\getDependencies{$p$}} {
    $D \gets \{ p \}$\;
    \While{{\normalfont \texttt{true}}}{
      $D' \gets \emptyset$\;
      \For{$d \in D$}{
        \For{$i \gets 1$ \KwTo $n$}{
          $\edgeExists \gets \mathbf{A}[i][d.\predicate] = \{ 1 \}$\;
          \uIf{$\edgeExists$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ i \}$\;
          }
          \uElseIf{$\edgeExists$ {\bf and not} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, d.\source, d.\target) \}$\;
          }
          \ElseIf{$|\mathbf{A}[i][d.\predicate]| > 1$ {\bf and} $d$.\isDetermined{}}{
            $D' \gets D' \cup \{ (i, i, d.\predicate) \}$\;
          }
        }
      }
      \lIf{$D' = D$}{\Return{$D$}}
      $D \gets D'$\;
    }
  }
  \caption{Computing the dependencies of a predicate}
\end{algorithm}

\section{Entailment Checking for Negative/All Cycles}
% TODO: explain that the hasNegativeCycles algorithm is an extension of a simple
% cycle detection algorithm.
% TODO: the difficulty with propagation is that there is no good way to define
% the adjacency matrix with 0/-1/+1 values, the negativity condition is too
% difficult.
% TODO: NB: the set of decision variables for this constraint must exclude all
% arguments.

\begin{algorithm}[t]
  \KwData{a constraint model for a logic program $\mathscr{P}$ amid execution}
  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
  $\mathscr{P}$ with all bodies and all predicates in heads fully
  determined\footnotemark\;
  \lIf{$\mathscr{R} = \emptyset$}{\Return{\textsc{undefined}}}
  \lIf{\hasNegativeCycles($G_{\mathscr{R}}$)}{\Return{\textsc{false}}}
  \lIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment checking for negative cycles}
\end{algorithm}
\footnotetext{Due to the definition of independence, the arguments of the head
  atom need not be determined.}

% \section{Conditional Independence}

% Finish the propagation algorithm for conditional independence. The
%  propagation algorithm for independence is extended with masks that are either
%  potential or definite. Masking happens in two stages: first, we mask
%  expressions within formulas, and then predicates. Masking algorithm uses an
%  algorithm for perfect bipartite matching.

% Both determined $\implies$ fail(). Both determined but at least is one masked by
% a (probable/determined) mask $\implies$ nothing. One determined $\implies$ the
% other one cannot exist.

% \begin{algorithm}
%   \SetKwProg{Fn}{Function}{:}{}
%   \SetKwFunction{potentialRoots}{potentialRoots}
%   \SetKwFunction{getTreeValues}{getTreeValues}
%   \SetKwFunction{getTreeStructureDomainValues}{getTreeStructureDomainValues}
%   \SetKwData{clause}{clause}
%   \KwData{connective $c$, a set of predicates $P$}
%   \Fn{\potentialRoots{\clause, $i$}} {
%     $R \gets \emptyset$\;
%     $V \gets \clause.\getTreeValues{$i$}$\;
%     \For{$v \in V$}{
%       \uIf{$v = c$}{
%         $R \gets R \cup \{ (i, |V| = 1) \}$\;
%       }
%       \ElseIf{$v \in P$}{
%         $R' \gets \clause.\getTreeStructureDomainValues{$i$}$\;
%         $R \gets R \cup \{ (r, |R'| = 1) \mid r \in R' \}$\;
%       }
%     }
%     \Return{$R$}\;
%   }
%   \caption{Potential root nodes of the required expression, assuming that the
%     node at index $i$ is part of the expression}
% \end{algorithm}

>>>>

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}

@@

\newcommand{\variable}[1]{\texttt{\textup{#1}}}
\newcommand{\arrayd}[3]{\variable{{#1}[}{#2}\variable{]} \in {#3}}
\newcommand{\arrayt}[3]{\variable{{#3}[}{#2}\variable{] {#1}}}
\newcommand{\logical}[1]{{\normalfont \texttt{#1}}}

\newcommand{\predicates}{\mathcal{P}}
\newcommand{\variables}{\mathcal{V}}
\newcommand{\constants}{\mathcal{C}}
\newcommand{\tokens}{\mathcal{T}}
\newcommand{\arities}{\mathcal{A}}
\newcommand{\maxArity}{\mathcal{M}_{\mathcal{A}}}
\newcommand{\maxNumNodes}{\mathcal{M}_{\mathcal{N}}}
\newcommand{\maxNumClauses}{\mathcal{M}_{\mathcal{C}}}

\DeclareMathOperator{\Determined}{Determined}
\DeclareMathOperator{\Undetermined}{Undetermined}
\DeclareMathOperator{\AlmostDetermined}{AlmostDetermined}

@@

\begin{constraint}[Redundant]
  For $v \in \variables{}$,
  \[
    \variable{introductions}[v] \ne 0 \quad \implies \quad
    \variable{introductions}[v] - 1 \in \variable{occurrences}[v].
  \]
\end{constraint}

@@

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$, \logical{false}} \times
      \getDependencies{$p_2$, \logical{false}}$ such that $d_1.\predicate =
      d_2.\predicate$}{
    \lIf{$d_1$ {\bf is} $\Determined$ {\bf and} $d_2$ {\bf is}
      $\Determined$}{\fail{}}
    \If{$d_1$ {\bf is} $\Determined$ {\bf and} $d_2$ {\bf is}
      $\AlmostDetermined(\_, s, t)$ {\bf or} \hspace{50pt} $d_2$ {\bf is} $\Determined$ {\bf
        and} $d_1$ {\bf is} $\AlmostDetermined(\_, s, t)$}{
      $\mathbf{A}[s][t]$.\removeValue{$1$}\;
    }
  }
  \caption{Propagation algorithm for predicate independence}
\end{algorithm}

\begin{algorithm}[p]
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwData{predicate}{predicate}
  \KwData{predicates $p_1$, $p_2$}
  $D \gets \{ (d_1, d_2) \in \getDependencies{$p_1$, \logical{true}} \times
  \getDependencies{$p_2$, \logical{true}} \mid d_1.\predicate = d_2.\predicate
  \}$\;
  \lIf{$D = \emptyset$}{\Return{\textsc{true}}}
  \lIf{$\exists (\Determined \_, \Determined \_) \in D$}{\Return{\textsc{false}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment check for predicate independence}
\end{algorithm}

% TODO: introduce this sum data type syntax
\begin{grammar}
  <dependency> ::= $\Determined p$ | $\Undetermined p$ | $\AlmostDetermined(p, s, t)$
\end{grammar}
where $p,s,t \in \predicates{}$. For a dependency $d$---regardless of its exact
type---we will refer to its predicate $p$ as $d.\variable{predicate}$.
% TODO: underscore represents a variable that can take any value (different
% underscores can take different values)

\begin{algorithm}[p]
  \SetKwData{edgeExists}{edgeExists}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \SetKwData{all}{allDependencies}
  \SetKwFunction{isDetermined}{isDetermined}
  \SetKwFunction{getDependencies}{getDependencies}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{a $|\predicates{}| \times |\predicates{}|$ adjacency matrix
    $\mathbf{A}$ with $\{ 0, 1 \}$ as the domain}
  \Fn{\getDependencies{$p$, \all}} {
    $D \gets \{ \Determined p \}$\;
    \While{\logical{true}}{
      $D' \gets \emptyset$\;
      \For{$d \in D$ {\bf and} $q \in \predicates{}$}{
        $\edgeExists \gets \mathbf{A}[q][d.\predicate] = \{ 1 \}$\;
        \uIf{$\edgeExists$ {\bf and} $d$ {\bf is} $\Determined$}{
          $D' \gets D' \cup \{ \Determined q \}$
        }
        \uElseIf{$\edgeExists$ {\bf and} $d$ {\bf is} $\AlmostDetermined(\_, s, t)$}{
          $D' \gets D' \cup \{ \AlmostDetermined(q, s, t) \}$\;
        }
        \uElseIf{$|\mathbf{A}[q][d.\predicate]| > 1$ {\bf and} $d$ {\bf is}
          $\Determined r$}{
          $D' \gets D' \cup \{ \AlmostDetermined(q, q, r) \}$\;
        }
        \ElseIf{$|\mathbf{A}[q][d.\predicate]| > 1$ {\bf and} \all}{
          $D' \gets D' \cup \{ \Undetermined q \}$\;
        }
      }
      \lIf{$D' = D$}{\Return{$D$}}
      $D \gets D'$\;
    }
  }
  \caption{Computing the dependencies of a predicate}
\end{algorithm}
% TODO: any way to test all of this?

% TODO: Discuss that the algorithms work in case there are multiple 'paths'
% without any issues.
% TODO: An example of algorithms, showing a program that's not fully determined
% yet, a dependency graph (with dashed and grayed out edges). Create a situation
% with multiple different dependency paths such that propagation would result in
% a solution).

\section{Entailment Checking for Negative/All Cycles}
% TODO: explain that the hasNegativeCycles algorithm is an extension of a simple
% cycle detection algorithm.
% TODO: the difficulty with propagation is that there is no good way to define
% the adjacency matrix with 0/-1/+1 values, the negativity condition is too
% difficult.
% TODO: NB: the set of decision variables for this constraint must exclude all
% arguments.
% In order to avoid thrashing, we fix the following variable ordering: [...]. Example of thrashing: P(X, a) <- ~P(X, b) contradicts the negative cycle constrant. Let's try P(X, a) <- P(X, c) instead (and so on).

\begin{algorithm}[t]
  \KwData{a constraint model for a logic program $\mathscr{P}$ amid execution}
  \SetKwFunction{hasNegativeCycles}{hasNegativeCycles}
  Let $\mathscr{R} \subseteq \mathscr{P}$ be the largest subprogram of
  $\mathscr{P}$ with all bodies and all predicates in heads fully
  determined\footnotemark\;
  \lIf{$\mathscr{R} = \emptyset$}{\Return{\textsc{undefined}}}
  \lIf{\hasNegativeCycles($G_{\mathscr{R}}$)}{\Return{\textsc{false}}}
  \lIf{$\mathscr{R} = \mathscr{P}$}{\Return{\textsc{true}}}
  \Return{\textsc{undefined}}\;
  \caption{Entailment check for negative cycles}
\end{algorithm}
\footnotetext{Due to the definition of independence, the arguments of the head
  atom need not be determined.}
